---
title: "Data cleaning"
format: html
---

```{r}
library(tidyverse)
library(here)

# Read in data
# Sampling site descriptions
sites <- read_csv(here("data", "knb-lter-cap.562.10", "300_sites_1.csv"))

# Absence or presence of mycorrhiza species
species_occurrences <- read_csv(here("data", "knb-lter-cap.562.10", "300_species_occurrance_1.csv"))

# Colonization of plant roots
colonization <- read_csv(here("data", "knb-lter-cap.562.10", "300_summary_data_1.csv"))
```
```{r}
# View data types
glimpse(sites)
```
I only need one primary key column in this data frame to connect to my other dataframes, site_id. Therefore, I am going to drop the generated ID column. 

It looks like all of the data types are as expected (character for site_id and site_description, datetime for sampled_on, and double for lat and long). 

```{r}
sites_clean <- sites %>% 
  select(-ID)
```

Next, we have species occurrences, which is the absence or presence of mycorrhiza species. This data frame has a lot of information, including a plant_id and a lot of information about each plant before it includes the absence or presence of these species. Based on this, I want to normalize the data further and create an intermediary table for just plants. 

This table will include a unique ID as the primary key (the existing plant_id is not unique, it repeats 1-10 for each site and plant type- for each site there are 10 succulent plants and 10 woody plants) to associate that plant with the other tables rows. 

```{r}
glimpse(species_occurrences)
```
```{r}
species_occurrences <- species_occurrences %>% 
  # add unique primary key 
   mutate(id = row_number()) %>% 
   relocate(id) %>% 
  # rename plant_id and id column 
  rename(plant_number = plant_id) %>% 
  rename(plant_id = id)
```



```{r}
# Select rows for intermediate table
plants_clean <- species_occurrences %>% 
  select(c(plant_id, site_id, plant_number, plant_type, plant_species)) 
```

Now that the intermediary plant table has been created, we can check the types.

```{r}
glimpse(plants_clean)
```
All of the types are as expected. 

Next, we can edit the species_occurrences table and replace the plant info with just the plant_id. 

```{r}
species_occurrences <- species_occurrences %>% 
  select(-c(ID, site_id, plant_number, plant_type, plant_species))
```

Now, we want to pivot the presence/absence columns from wide to long format, with a mychorrhiza_species column and a presence column (0/1).

```{r}
species_occurrences_clean <- species_occurrences %>% 
  pivot_longer(
    cols = acaulospora_delicata:acaulospora_mellea,
         names_to = "mycorrhiza_species",
         values_to = "present")
```

```{r}
glimpse(species_occurrences_clean)
```
The types look good!

Let's look at colonization 
```{r}
glimpse(colonization) 
```
We can similarly replace the plant info columns with a plant id, and we don't really need the id. 

```{r}
colonization <- colonization %>%
  # rename same columns to match for joining 
  rename(plant_number = plant_id) %>%
  # left join to match correct plant_id to row
  left_join(plants_clean, by = c("site_id", "plant_number", "plant_type", "plant_species")) %>%
  # remove redundant columns
  select(-c(ID, site_id, plant_number, plant_type, plant_species)) %>% 
  # move plant_id to the front 
  relocate(plant_id)
```

```{r}
glimpse(colonization)
```
The types for each column in our cleaned colonization df match the metadata!
There are a few rows with all NA's at the bottom of this data set, so we can remove them. 

```{r}
# remove rows with all NAs
colonization_clean <- colonization %>%
  filter(!is.na(plant_id)) %>% 
  # remove row where all other columns are NA
  filter(!if_all(-plant_id, is.na)) 
```

Now that the data is cleaned, the types of each column are checked, and normalized, we can write them to new csv files for ingestion into duckdb. 

```{r}
# file names
datadir_processed <- file.path("processed_data")

# check if the folder exists
dir.create(datadir_processed, showWarnings = FALSE)

# write the files
write_csv(sites_clean, file.path(datadir_processed, "sites_clean.csv"))

write_csv(plants_clean, file.path(datadir_processed, "plants_clean.csv"))

write_csv(species_occurrences_clean, file.path(datadir_processed, "species_occurrences_clean.csv"))

write_csv(colonization_clean, file.path(datadir_processed, "colonization_clean.csv"), na = "")
```

# clean data
# write to csv to root directory, series of tables
# upload csvs to repo
# create database folder
# open csv in terminal 
# Create database- D duckdb newdatabase.db
